Active Learning from Blackbox to Timed Connectors

### 2016-01-17 paper/paper.tex
.

\begin{document}
\maketitle
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized mealy machine as an operational semantic
  model for channel-based coordination language Reo. Parameterized mealy machine serves as a bridge
  between Reo connectors and mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and transformed into mealy machines. Moreover, we adapt L*,
  a well-known learning algorithm, to timed connectors (in the form of mealy machines). The new
  algorithm has shown its efficiency in multiple case studies. 
  Implementations of this framework is provided as a package in \texttt{Golang}.
\end{abstract}

@@

Distributed real-time embedded system (DRES) is reforming our lives with the name \emph{IoT}, the
internet of things, wherein systems are usually composed of individual components and a
\emph{connector}. Such systems could be distributed logically or physically, which
makes coordination processes even more complicated. In this case, we need to specify these coordination
processes with so-called \emph{coordination languages}, so that formal techniques can be applied to
guarantee their reliability.

@@

Formal verification techniques have also been proved applicable in timed connectors, e.g.
\cite{DBLP:conf/tase/LiCWS15} has shown us a comformance testing method on timed connectors.
Bounded model checking methods based on SAT solvers were also adapted in
\cite{DBLP:journals/scp/Kemper12}.
All these solutions seem practical and impressive, however, a common question is faced by
most of them: \emph{how to obtain these formal models?}

\emph{Correctness} of connectors is very much related to some low-level implementation details.
For example, well-writen code may behave dramatically weird with an improper set of concurrency
primitives. Such things happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually modeling an existing connector seems rather unreliable,
even with reference to its source code.

@@

To address the problem, we proposed a active learning framework (shown in Figure \ref{fig:howto}) to
automatically extract timed connectors from blackbox models. To achieve this goal, firstly we
present \emph{parameterized Mealy machine} as a parametierized semantics for timed Reo channels,
which is able to generate concrete Mealy machines with a given alphabet. Then the $L*$ algorithm
\cite{DBLP:journals/iandc/Angluin87}
is adapted and optimized to extract Mealy machines with time action from the \emph{connectors in
blackboxes}.

The rest of the paper is organized as follows. After this general introduction, in Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including Reo the coordination
language, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines an
operational semantics of Reo, which is used in Section \ref{sec:activelearning} to show how to
extract Reo models from blackboxes by means of active learning. Finally, in Section
\ref{sec:experiment}, we discuss the implementation and optimization in \texttt{Golang}.

@@

Reo is a channel based exogenous coordination language proposed by F. Arbab in
\cite{DBLP:journals/mscs/Arbab04}. 
A Reo model, also called \emph{connector}, provides the protocol
that formalizes the communication, synchronization and cooperation among the components which
communicate through the connector. Connectors can be defined without dependence on components,
which makes Reo a powerful ``glue language'' in component-based
development\cite{DBLP:journals/sigsoft/Gill03}.

In Reo, complex connectors are made up of simpler ones. The atomic connectors are called
\emph{channels}, and each of them has two \emph{channel ends}. There are two type of channel ends:
\emph{source} and \emph{sink}. Source channel ends accepts data into the channel, while sink
channels ends dispense them out of the channel.

@@

The behavior of some channels are informally described as follows. (Graphical representations
can be found in Figure \ref{fig:basic}).

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff the data item can be
    dispensed to its sink end simultaneously.
  \item [-] A \emph{LossySync} channel is always prepared to accept data items. These items will be
    send to its sink end simutaneously if possible, otherwise they will be dropped.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff a data item is also available to accept simultaneously
    through the other end. Then both items will be dropped.
  \item [-] A \emph{FIFO1} channel is an asychronous channel with a buffer cell. It accepts a
    data item whenever the buffer is empty. 
\end{itemize}

Channels are attached on component instances or \emph{nodes}. There are three types of nodes:
\emph{source}, \emph{sink} and \emph{mixed node}, depending on whether all channel ends that
coincide on a node are source ends, sink ends or both. (see in Figure \ref{fig:node})

@@

With definition of more basic channels, it's easy to extend Reo to formalize coordination in
different areas. In this paper, we take timed Reo\cite{DBLP:conf/sefm/ArbabBBR04} as our formal
model. Timed Reo includes several timed channels, where the most commonly-used one, called
\emph{timer}, is also shown in Figure \ref{fig:basic}. A t-timer channel accepts any data item from
its source end, and later dispense it to its sink end after a delay of t time units.

@@

As shown in Figure \ref{fig:reoconnector}, we use a simple example to illustrate how complex
connectors are constructed and used in coordination. In this example, COM1,COM2 and
MONITOR are components. With an \emph{alternator} connector, MONITOR can obtain data items from
COM1 and COM2 alternately. The \emph{alternator} example has been implemented in \texttt{Golang} as
shown in Section \ref{sec:reolib}.
Besides, all the connectors are supposed to be \emph{deterministic} hereinafter to meet the
requirement of active learning algorithm L*.

@@

In this work, we use \emph{Mealy machines} to model Reo connectors.
As an extension of \emph{finite state machine}, Mealy machine was first proposed by George. H. Mealy
in \cite{George1955A}. Compared with other variants, Mealy machines are designed to model
reactive systems, where outputs are determined not only by its current state but also the current
inputs. Besides, Mealy machines are supposed to be \emph{input enabled}, which means that all
possible inputs should be acceptable in all states. In other words, if an input is invalid for some
state, we need to manually define an additional state to describe such exceptions.

Various forms of Mealy machines are defined in different works. 
Since active automata learning very much depends on the system-under-learn to be deterministic, in
this paper we formally define a deterministic Mealy machine following
\cite{DBLP:conf/sfm/SteffenHM11}.

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, \delta, \lambda)$ consisting of the following:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] a start state (also called initial state) $s_0$ which is an element of $S$
    \item[-] a finite set called the input alphabet $I$
    \item[-] a finite set called the output alphabet $O$
    \item[-] a transition function $\delta : S \times I \rightarrow S$ mapping pairs of a
      state and an input symbol to the corresponding next state.
    \item[-] an output function $\lambda : S \times I \rightarrow O$ mapping pairs
      of a state and an input symbol to the corresponding output symbol.
  \end{itemize}
\end{definition}

We say that a Mealy machine is \emph{finite} if the set $S$ of states and the set $I$ of inputs are
finite. Hereinafter, Mealy machines are always supposed to be finite.

@@

Figure \ref{fig:activelearning} shows the sketch of active learning, wherein:
\begin{itemize}
  \item[-] \emph{Teacher} and \emph{Student}: Active learning is an interactive process where
    students ask questions and teachers answer. Here learning algorithm plays the role of student.
  \item[-] \emph{Oracles} is an interface specifying which kind of questions can be answered by the
    teacher.
  \item[-] \emph{SUL} is an abbreviation of System Under Learn. The name comes from a well-known
    concept SUT (System Under Test) in software testing. In this case, we take blackbox models as our
    SULs.
  \item[-] \emph{CEGAR} indicates Counter-Example Guided Abstraction
    Refinement\cite{DBLP:conf/cav/ClarkeGJLV00}. In active learning, we need counter-examples to
    guide us on further quries and cover the undistinguished states.
\end{itemize}

When applying active automata learning on some model, firstly we assume that it should be equivalent
to some \emph{Mealy machine}, in other words, a deterministic model accepting a finite set of input
symbols and mapping them to a finite set of output symbols.

@@

\begin{itemize}
  \item[-] \textbf{Membership Query} (hereinafter referred to as \emph{mq}) The name comes from
    some grammar-learning papers (e.g. \cite{DBLP:journals/iandc/Angluin87}), where \emph{mq} checks
    if a word is a member of certain language defined by the given grammar. When it comes to
    automata learning, \emph{mq} is supposed to provide \emph{simulation results} for given input
    sequences.
  \item[-] \textbf{Equivalence Query} (hereinaftr referred to as \emph{eq}) Given a hypothesis
    (usually constructed by the learning algorithm), \emph{eq} checks whether the hypothesis is
    equivalent to the system-under-learn and generates a counter-example if needed. Generally, 
    \emph{equivalence query} is irrealizable when SUL is a blackbox. So we tactically use membership
    queries to achieve the approximate results.
\end{itemize}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}
In this section, we illustrate how to transform a timed connector into a Mealy machine with time
action.
Since time is not involed in original Mealy machine, we first discuss how to formalize the time
dimension in Mealy machine. After that, we present a \emph{parameterized Mealy
machine} as a bridge between connectors and Mealy machines.

\subsection{Time Domain}
Time is involved in several extensions of Reo. For example, timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04}, hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined in
$\mathbb{R}$. However, we also found some works like \cite{DBLP:journals/fmsd/PrabhakarDM015} where
rational time indeed  makes things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behaviors greatly.

As presented in section \ref{sec:reo}, all real-time behavior in timed Reo comes with the
\emph{t-timer} channels, and the number of these channels are apparently finite. We use
$t_i\in\mathbb{Q}$ to denote the delays of these timer channels, and now we can define a precision
function $prec$.
\[
prec(t_1,\cdots,t_n) = \max_T\{\forall t_i.\exists n_i\in\mathbb{N}.t_i=n_i\cdot T\}
\]
It's easy to prove that such a $T$ is always existing.

@@

Besides, we're going to add a ``T'' action in Mealy machines. It indicates that a transition
will take a time unit to finish.

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized Mealy machine} (hereinafter referred to as PMM)
to model timed connectors. PMM is supposed to behave as a middle representation between Mealy
machines and timed connectors. Connectors are firstly defined as parameterized Mealy machines, and
composed by production operator and link operator. Then, concrete Mealy-machine model can be
generated from the PMM model.
Following the formal definition of Mealy machine in Section \ref{sec:preliminaries}, we define the
\emph{parameterized Mealy machine} as follows.

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized Mealy machine(PMM) as an operational semantic
  model for channel-based coordination language Reo. Parameterized Mealy machine serves as a bridge
  between Reo connectors and mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and transformed into mealy machines. Moreover, we adapt L*,
  a well-known learning algorithm, to timed connectors (in the form of mealy machines). The new
  algorithm has shown its efficiency in multiple case studies. 
  Implementations of this framework is provided as a package in \texttt{Golang}.
\end{abstract}

@@

Distributed real-time embedded systems (DRES) are reforming our lives with the
Internet of Things(IoT), wherein individual components are composed via a
\emph{connector} to build a system. Such systems could be distributed logically or physically, which
makes coordination processes even more complicated. In this case, we need to specify the coordination
processes with \emph{coordination languages}, such that formal techniques can be applied to
guarantee their reliability.

@@

Formal verification and validation techniques have also been proved applicable for timed connectors, 
e.g. a comformance testing framework on timed connectors in \cite{DBLP:conf/tase/LiCWS15}.
A SAT-based bounded model checking approach was adapted in \cite{DBLP:journals/scp/Kemper12} for
verifying connectors. Although these solutions are practical and impressive, there is one important
question faced by most of them: \emph{how to obtain the formal models?}

\emph{Correctness} of connectors is closely related to some low-level implementation details.
For example, well-written code may behave dramatically weird with an improper set of concurrency
primitives. Such senarios happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually modeling an existing connector seems rather unreliable,
even with reference to its source code.

@@

In this paper we address this question by proposing a active learning framework as shown in Figure
\ref{fig:howto}) to automatically extract timed connector models from blackbox implementations. 
To achieve this goal, we first introduce \emph{parameterized Mealy machine} as a parametierized
semantics for timed Reo connectors, which can be transformed intoe concrete Mealy machines with a
given alphabet. Then the $L*$ algorithm \cite{DBLP:journals/iandc/Angluin87} is adapted and
optimized to extract Mealy machines with timed action from the connector implementations as
blackboxes.

The rest of the paper is organized as follows: In Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including the coordination
language Reo, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines an
operational semantics of Reo, which is used in Section \ref{sec:activelearning} to show how to
extract Reo models from blackboxes by means of active learning. In Section
\ref{sec:experiment}, we discuss the implementation and optimization of the approach in
\texttt{Golang}.

@@

Reo is a channel based exogenous coordination language proposed by F. Arbab in
\cite{DBLP:journals/mscs/Arbab04}. 
A coordinator in Reo, also called \emph{connector}, provides the protocol that controls and
organizes the communication, synchronization and cooperation among the components which
communicate through the connector. Connectors can be defined without dependence on components,
which makes Reo a powerful ``glue language'' in component-based
development\cite{DBLP:journals/sigsoft/Gill03}.

In Reo, complex connectors are made up of simpler ones. The atomic connectors are called
\emph{channels}, and each of them has two \emph{channel ends}. There are two types of channel ends:
\emph{source} and \emph{sink}. Source channel ends accept data into the channel, while sink
channel ends dispense data out of the channel.

@@

Figure \ref{fig:basic} shows the graphical representation of some basic channel types in Reo, whose
behavior are described as follows:

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff the data item can be
    dispensed through its sink end simultaneously.
  \item [-] A \emph{Lo ssySync} channel is always ready to accept data items. These items will be
    dispensed through its sink end if possible, otherwise they will be lost.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff a data item is also available to be accepted
    simultaneously through the other end. Then both items will be dropped.
  \item [-] A \emph{FIFO1} channel is an asychronous channel with a buffer cell. It accepts a
    data item whenever the buffer is empty. 
  \item [-] A t-timer channel accepts any data item from its source end, and later dispense it to
    its sink end after a delay of $t$ time units.

\end{itemize}

Channels are joint together in \emph{nodes}. There are three types of nodes:
\emph{source}, \emph{sink} and \emph{mixed} nodes, depending on whether all channel ends that
coincide on a node are source ends, sink ends or a combination of both.
(see in Figure \ref{fig:node})

@@

@@

Figure \ref{fig:reoconnector} provides a simple example to illustrate how complex
connectors are constructed and used in coordination. In this example, we have three components
COM1,COM2 and MONITOR. With an \emph{alternator} connector as shown in Figure
\ref{fig:reoconnector}, MONITOR can obtain data items from COM1 and COM2 alternately. This example
has been implemented in \texttt{Golang} as shown in Section \ref{sec:reolib}.
Besides, all the connectors are supposed to have \emph{deterministic} behavior hereinafter to meet
the requirement of active learning algorithm L*.

@@

In this work, we use \emph{Mealy machines} to model Reo connectors.
Mealy machine was first proposed by George. H. Mealy in \cite{George1955A} as an extension of finite
state machine. Compared with other variants, Mealy machines are designed to model
reactive systems, where outputs of a machine are determined by not only its current state but also
the current inputs. Besides, Mealy machines are supposed to be \emph{input enabled}, which means
that all possible inputs should be acceptable in all states. In other words, if an input is invalid
for some state, we need to manually define an additional state to describe such exceptions.

Various forms of Mealy machines have been defined in the literature. 
Since active automata learning requires the system-under-learn to be deterministic, in
this paper we take the deterministic Mealy machine model as defined in 
\cite{DBLP:conf/sfm/SteffenHM11}.

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, \delta, \lambda)$ consisting of:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] an initial $s_0\in S$
    \item[-] a finite set of inputs $I$
    \item[-] a finite set of outputs $O$
    \item[-] a transition function $\delta : S \times I \rightarrow S$ mapping a pair of a
      state and an input to the corresponding next state.
    \item[-] an output function $\lambda : S \times  I \rightarrow O$ mapping a pair
      of a state and an input to the corresponding output.
  \end{itemize}
\end{definition}

We say that a Mealy machine is \emph{finite} if the set $S$ of states and the set $I$ of inputs are
both finite. Hereinafter, Mealy machines are always supposed to be finite in this paper.

@@

Figure \ref{fig:activelearning} shows the sketch of active learning, wherein:
\begin{itemize}
  \item[-] \emph{Teacher} and \emph{Student}: Active learning is an interactive process where
    students ask questions and teachers answer. Here learning algorithm plays the role of student.
  \item[-] \emph{Oracles} is an interface specifying which kind of questions can be answered by the
    teacher.
  \item[-] \emph{SUL} is an abbreviation of System Under Learn. In this paper, we take blackbox
    models as our SULs.
  \item[-] \emph{CEGAR} indicates Counter-Example Guided Abstraction
    Refinement\cite{DBLP:conf/cav/ClarkeGJLV00}. In active learning, we need counter-examples to
    guide us on further queries and cover the undistinguished states.
\end{itemize}

When applying active automata learning on some model, we assume that the model should be equivalent
to some \emph{Mealy machine}, i.e., a deterministic model accepting a finite set of inputs and
mapping them to a finite set of outputs.

@@

\begin{itemize}
  \item[-] \textbf{Membership Query} (hereinafter referred to as \emph{mq}) In grammar-learning
    \cite{DBLP:journals/iandc/Angluin87}) \emph{mq} checks if a word is a member of certain language
    defined by the given grammar. When it comes to
    automata learning, \emph{mq} is supposed to provide \emph{simulation results} for given input
    sequences.
  \item[-] \textbf{Equivalence Query} (hereinaftr referred to as \emph{eq}) Given a hypothesis
    (usually constructed by the learning algorithm), \emph{eq} checks whether the hypothesis is
    equivalent to the system-under-learn and generates a counter-example if needed. Generally, 
    \emph{eq} is irrealizable when SUL is a blackbox. So we tactically use $mq$ to achieve the
    approximate results.
\end{itemize}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}
In this section, we illustrate how to transform a timed connector into a Mealy machine with timed
action.
Since time is not involved in original Mealy machine, we first discuss how to formalize the time
dimension in Mealy machine. After that, we present the notion of \emph{parameterized Mealy
machine} as a bridge between connectors and Mealy machines.

\subsection{Time Domain}
Time has been investigated in several extensions of Reo. For example, timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04, DBLP:conf/fmoods/MengA07}, hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined on the
real number field $\mathbb{R}$. However, there are also found some works like
\cite{DBLP:journals/fmsd/PrabhakarDM015} where
rational time indeed  makes things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behavior greatly.

As presented in section \ref{sec:reo}, all real-time behavior in timed Reo comes with a finite set
of timed channels, and the number of these channels are apparently finite. We use
$t_i\in\mathbb{Q}, i=1,2,\dots,n$ to denote the delays of these timed channels, and then we can define a precision
function $prec$:
\[
prec(t_1,\cdots,t_n) = \max_T\{T\in\mathbb{Q}|\forall t_i.\exists n_i\in\mathbb{N}.t_i=n_i\cdot T\}
\]
It's easy to prove that such a $T$ always exists.

@@

Besides, we're going to add a ``T'' action in Mealy machines. It indicates that the corresponding
transition will take one time unit to finish.

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized Mealy machine}to model timed connectors. PMM is
supposed to behave as a bridge between Mealy machines and timed connectors. Connectors are firstly
defined as PMMs, and composed by production operator and link operator. Then, concrete Mealy-machine
model can be generated from the PMM model.
Following the formal definition of Mealy machine in Section \ref{sec:preliminaries}, we define the
\emph{parameterized Mealy machine} as follows.
